---
title: "Overview of the DuckDBSpatial Package"
author:
- name: Patrick Aboyoun
  email: aboyounp@gene.com
  affiliation: Genentech, Inc.
date: "Compiled: `r format(Sys.time(), '%B %d, %Y')`"
package: DuckDBSpatial
vignette: >
  %\VignetteIndexEntry{Overview of the DuckDBSpatial Package}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
output:
  BiocStyle::html_document:
    number_sections: true
    toc: true
    toc_depth: 3
---

```{r setup, echo=FALSE}
library(BiocStyle)
knitr::opts_chunk$set(error=FALSE, warning=FALSE, message=FALSE)
```

# Introduction

The `r Biocpkg("DuckDBSpatial")` package extends `r Biocpkg("DuckDBDataFrame")`
with `sf`-compatible spatial methods on lazy DuckDB-backed columns and tables.
It supports GeoParquet 1.0 I/O and layer-level spatial predicates for point and
geometry workflows used in spatial transcriptomics pipelines.

# Lazy spatial columns

Bundled example data lives in `inst/extdata/spatial/` as partitioned Parquet.
Open it as a lazy `DuckDBDataFrame` and apply `sf` generics to the geometry
column without materializing the full table.

```{r}
library(DuckDBSpatial)
spatial_path <- system.file("extdata", "spatial", package = "DuckDBSpatial")
df <- DuckDBDataFrame(spatial_path)
df <- df[which(!is.na(df$type)), ]
geom <- df[["geometry"]]
head(st_geometry_type(geom))
head(st_area(geom))
```

Table-level helpers such as `st_filter()` work on the lazy table:

```{r}
query_pt <- sf::st_sfc(sf::st_point(c(30, 10)))
filtered <- st_filter(df, query_pt)
nrow(filtered)
```

# Layer-level spatial engines

For coordinate columns (for example transcript centroids), use the layer helpers
instead of a dedicated geometry column:

```{r}
pts_path <- tempfile(fileext = ".csv")
write.csv(data.frame(x = c(1, 5, 30), y = c(1, 5, 10)), pts_path, row.names = FALSE)
pts <- DuckDBDataFrame(pts_path, datacols = c("x", "y"))
poly <- sf::st_as_sfc("POLYGON((0 0, 6 0, 6 6, 0 6, 0 0))")
layerSpatialOverlaps(pts, poly, coords = c("x", "y"))
layerSubsetByGeometry(pts, poly, coords = c("x", "y"))
layerSpatialMatch(pts, df, coords = c("x", "y"))
unlink(pts_path)
```

# GeoParquet I/O

`readGeoParquet()` enables DuckDB GeoParquet conversion and returns a lazy
`DuckDBDataFrame`. `writeGeoParquet()` writes `sf` objects with GeoParquet 1.0
metadata (requires the Suggested `nanoparquet` package).

```{r}
ddb <- readGeoParquet(spatial_path)
nrow(ddb)
```

```{r eval=requireNamespace("nanoparquet", quietly=TRUE)}
pts <- sf::st_sf(id = 1:2, geometry = sf::st_sfc(sf::st_point(0:1), sf::st_point(2:3)))
path <- tempfile(fileext = ".parquet")
writeGeoParquet(pts, path)
readGeoParquet(path)
unlink(path)
```

# Session information

```{r}
sessionInfo()
```
